\bf Estimating the number of clusters using cross-validation

### 2019-06-29 JCGS-R3/appendix.tex
move cluster material to appendix

>>>>

\section{Enrichment analysis}

To further validate our clusters, we follow \citet{tavazoie1999systematic},
performing an enrichment analysis to discover which functional gene groups are
significantly over-represented in each cluster. In the Saccharomyces Genome
Database, each gene is mapped to a set of Gene Ontology categories. We focus
on the $103$ biological process categories.

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Biological process enrichment within gene clusters}
\footnotesize
\begin{tabular}{ccll}
\toprule
Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
\midrule
$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
\bottomrule
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {}
\end{table}

For each category and each cluster, we compute a $p$-value for the null
hypothesis that genes from the category are distributed across all clusters
without any bias towards the particular cluster in question. Under the null
hypothesis, the number of genes from the category that end up in the cluster
is distributed as a hypergeometric random variable. For each cluster, we
compute $p$-values for all $103$ biological process categories, and we report
those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
Bonferroni correction to control the family-wise error rate at level 5\%, we
only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.

From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
genes that somatize cell stress, such as oxidative heat-induce proteins.
Cluster~2 contains genes that govern mitochondrial translation and
mitochondrion organization. Cluster~3, the first period cluster, contains cell
cycle genes related to budding and cell polarity, along with genes that govern
RNA processing and transcription. Cluster~4 contains genes related to
cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
periodic cluster, contains genes that participate cell-cycle processes, along
with DNA replication and DNA repair.

\section{Comparison with Tavazoie~clusters}

In the \citet{tavazoie1999systematic} analysis, those authors performed
$k$-means clustering with $k = 30$; they found 23 of the clusters to be
uninterpretable, and they found 7 clusters to be meaningful. To compare our
clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
comparing our clusters with the 7 interpretable Tavazoie clusters in
Table~\ref{table:confusion}.
%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
%% Tavazoie clusters, including the non-interpretable ones.)
Entry $(i,j)$ of the confusion matrix gives the number of genes in
Tavazoie's Cluster~$i$ and our Cluster~$j$.

\begin{table}
\caption{\label{table:confusion} Confusion matrix comparing
    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
    clusters found by the \cite{tavazoie1999systematic} analysis}
\centering
\begin{tabular}{lcccccr}
  \toprule
 & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
  \midrule
Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
  \midrule
  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
   \bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
$(i,j)$ of the upper left part of this figure gives the mean expression level
for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
the plots in the margins give the mean expression levels for Tavazoie's
clusters (top right) and our clusters (bottom left). In
Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
number of genes in that cell is greater than $20$.

Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
uninterpretable clusters. Our Cluster~2 contains high concentrations of
Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
highlited these clusters as being periodic.

For the clusters that Tavazoie et al.\ were able to characterize, our analysis
broadly agrees with the earlier clustering. The major difference between our
analysis and that of \citet{tavazoie1999systematic} is that we are able to
identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
instead of $k = 30$), and we are able to interpret all of the clusters
found by our analysis.

### 2019-06-29 JCGS-R3/cvclust.tex
move cluster material to appendix

\subsection{Enrichment analysis}

To further validate our clusters, we follow \citet{tavazoie1999systematic},
performing an enrichment analysis to discover which functional gene groups are
significantly over-represented in each cluster. In the Saccharomyces Genome
Database, each gene is mapped to a set of Gene Ontology categories. We focus
on the $103$ biological process categories.

\begin{table}
\begin{center}
\captionsetup{justification=centering}
\caption{\label{table:enrich} Biological process enrichment within gene clusters}
\footnotesize
\begin{tabular}{ccll}
\toprule
Cluster & Cluster Size & Process Category (In Cluster/Total Genes) & $p$-value \\
\midrule
$1$ & $550$ & response to oxidative stress ($24/55$)  & $1.5\times10^{-5}$  \\
    &       & response to chemical ($64/213$)         & $2.2\times10^{-5}$  \\
$2$ & $590$ & mitochondrion organization ($79/159$)   & $1.1\times10^{-16}$ \\
    &       & mitochondrial translation ($28/51$)     & $2.9\times10^{-8}$  \\
    &       & generation of precursor metabolites and energy ($37/80$) & $7.3\times10^{-8}$ \\
$3$ & $654$ & transcription from RNA polymerase II promoter ($75/214$) & $5.5\times10^{-6}$ \\
    &       & mRNA processing ($30/67$)               & $2.7\times10^{-5}$  \\
    &       & mitotic cell cycle ($63/183$)           & $6.2\times10^{-5}$  \\
$4$ & $634$ & cytoplasmic translation ($105/134$)     & $3.3\times10^{-47}$ \\
    &       & ribosomal subunit biogenesis ($73/138$) & $7.7\times10^{-17}$ \\
    &       & rRNA processing ($61/131$)              & $5.7\times10^{-11}$ \\
    &       & ribosome assembly ($21/36$)             & $1.5\times10^{-6}$  \\
$5$ & $517$ & chromosome segregation ($53/106$)       & $6.0\times10^{-15}$ \\
    &       & cellular response to DNA damage stimulus ($71/172$)      & $3.6\times10^{-14}$ \\
    &       & DNA repair ($64/147$)                   & $3.7\times10^{-14}$ \\
    &       & DNA replication ($42/78$)               & $1.8\times10^{-13}$ \\
    &       & mitotic cell cycle ($70/183$)           & $4.8\times10^{-12}$ \\
\bottomrule
\end{tabular}
\end{center}
\hspace{0.5in} \footnotesize {}
\end{table}

For each category and each cluster, we compute a $p$-value for the null
hypothesis that genes from the category are distributed across all clusters
without any bias towards the particular cluster in question. Under the null
hypothesis, the number of genes from the category that end up in the cluster
is distributed as a hypergeometric random variable. For each cluster, we
compute $p$-values for all $103$ biological process categories, and we report
those that are significantly enrigched in Table~\ref{table:enrich}.  Using a
Bonferroni correction to control the family-wise error rate at level 5\%, we
only report $p$-values that are less than $0.05 / 103 = 4.8\times10^{-4}$.

From Table \ref{table:enrich}, we can see that Cluster~1 is enriched with
genes that somatize cell stress, such as oxidative heat-induce proteins.
Cluster~2 contains genes that govern mitochondrial translation and
mitochondrion organization. Cluster~3, the first period cluster, contains cell
cycle genes related to budding and cell polarity, along with genes that govern
RNA processing and transcription. Cluster~4 contains genes related to
cytoplasmic translation and genes encoding ribosomes. Cluster~5, the second
periodic cluster, contains genes that participate cell-cycle processes, along
with DNA replication and DNA repair.

\subsection{Comparison with Tavazoie~clusters}

In the \citet{tavazoie1999systematic} analysis, those authors performed
$k$-means clustering with $k = 30$; they found 23 of the clusters to be
uninterpretable, and they found 7 clusters to be meaningful. To compare our
clusters with the Tavazoie et al.\ clusters, we prepared a confusion matrix
comparing our clusters with the 7 interpretable Tavazoie clusters in
Table~\ref{table:confusion}.
%%(Appendix~\ref{app:confusion} contains a confusion matrix for all 30
%% Tavazoie clusters, including the non-interpretable ones.)
Entry $(i,j)$ of the confusion matrix gives the number of genes in
Tavazoie's Cluster~$i$ and our Cluster~$j$.

\begin{table}
\caption{\label{table:confusion} Confusion matrix comparing
    the 5 clusters found by Gabriel cross-validation to the 7 interpretable
    clusters found by the \cite{tavazoie1999systematic} analysis}
\centering
\begin{tabular}{lcccccr}
  \toprule
 & Cluster 1 & Cluster 2 & Cluster 3 & Cluster 4 & Cluster 5 & Total \\ 
  \midrule
Cluster 1 & 0 & 0 & 1 & 161 & 2 & 164 \\ 
  Cluster 2 & 1 & 0 & 0 & 0 & 185 & 186 \\ 
  Cluster 3 & 0 & 0 & 91 & 11 & 2 & 104 \\ 
  Cluster 4 & 0 & 102 & 2 & 66 & 0 & 170 \\ 
  Cluster 7 & 1 & 10 & 83 & 7 & 0 & 101 \\ 
  Cluster 8 & 3 & 145 & 0 & 0 & 0 & 148 \\ 
  Cluster 14 & 0 & 1 & 29 & 6 & 38 & 74 \\ 
  Other & 545 & 332 & 448 & 383 & 290 & 1998 \\ 
  \midrule
  Total & 550 & 590 & 654 & 634 & 517 & 2945 \\ 
   \bottomrule
\end{tabular}
\end{table}

Figure~\ref{fig:5_clusters} provides a more in-depth comparison with the
Tavazoie clusters, using a graphical confusion matrix.  The plot in cell
$(i,j)$ of the upper left part of this figure gives the mean expression level
for genes in the intersection of Tavazoie's Cluster~$i$ and our Cluster~$j$;
the plots in the margins give the mean expression levels for Tavazoie's
clusters (top right) and our clusters (bottom left). In
Figure~\ref{fig:5_clusters}, we only include a plot for cell $(i,j)$ if the
number of genes in that cell is greater than $20$.

Our Cluster~1 mainly consists of genes that Tavazoie et al.\ found to be in
uninterpretable clusters. Our Cluster~2 contains high concentrations of
Tavazoie's Clusters~4 and~8. Our first periodic cluster, Cluster~3, contains
high concentrations of Tavazoie's Clulsters~3,~7, and~14; this is notable,
because Tavazoie et al.\ highlighted their Clusters~7 and~14 as being
periodic. Our Cluster~4 contains almost all of Tavazoie's Cluster~1, along
with part of Tavazoie's Cluster~4. Finally, our second periodic cluster,
Cluster~5, contains almost all of Tavazoie's Cluster~2, along with part of
Tavazoie's Cluster 14; this, again, is notable, because Tavazoie et al.\
highlited these clusters as being periodic.

For the clusters that Tavazoie et al.\ were able to characterize, our analysis
broadly agrees with the earlier clustering. The major difference between our
analysis and that of \citet{tavazoie1999systematic} is that we are able to
identify meaningful groups of genes with a much smaller value of $k$ ($k = 5$
instead of $k = 30$), and we are able to interpret all of the clusters
found by our analysis.

>>>>

%% TODO: enrichment analysis

%% TODO: mention comparison with Tavazoie clusters
